---
title: "Danish Kings"
author: "Luna Marcher"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The task here is to load your Danish Monarchs csv into R using the `tidyverse` toolkit, calculate and explore the kings' duration of reign with pipes `%>%` in `dplyr` and plot it over time.

## Load the kings

Make sure to first create an  `.Rproj` workspace with a `data/` folder where you place either your own dataset or the provided `kings.csv` dataset. 

1. Look at the dataset that are you loading and check what its columns are separated by? (hint: open it in plain text editor to see)

List what is the

separator: comma 

2. Create a `kings` object in R with the different functions below and inspect the different outputs.
-  `read.csv()`
-  `read_csv()`
-  `read.csv2()` 
-  `read_csv2()`
```{r eval=TRUE}
# FILL IN THE CODE BELOW and review the outputs
library(tidyverse)
kings1 <- read.csv("data/danish_monarchs_clean.csv")

kings2 <- read_csv("data/danish_monarchs_clean.csv")

kings3 <- read.csv2("data/danish_monarchs_clean.csv")

kings4 <- read_csv2("data/danish_monarchs_clean.csv")
```
Answer: 
1. Which of these functions is a `tidyverse` function? Read data with it below into a `kings` object
(the fourth one because my data is separated by ";")
2. What is the result of running `class()` on the `kings` object created with a tidyverse function.
(it confirms the object is a tibble)
3. How many columns does the object have when created with these different functions?
(It has 5 columns)
4. Show the dataset so that we can see how R interprets each column
```{r eval = TRUE}
# COMPLETE THE BLANKS BELOW WITH YOUR CODE, then turn the 'eval' flag in this chunk to TRUE.

kings <- read_csv2("data/danish_monarchs_clean.csv")

class(kings)


head(kings)


glimpse(kings)


```


## Calculate the duration of reign for all the kings in your table
You can calculate the duration of reign in years with `mutate` function by subtracting the equivalents of your `startReign` from `endReign` columns and writing the result to a new column called `duration`. But first you need to check a few things:

* Is your data messy? Fix it before re-importing to R
* Do your start and end of reign columns contain NAs? Choose the right strategy to deal with them: `na.omit()`, `na.rm=TRUE`, `!is.na()`


Create a new column called  `duration` in the kings dataset, utilizing the `mutate()` function from tidyverse. Check with your group to brainstorm the options.

```{r eval = TRUE}
view(kings) 

kings <- kings %>%
  mutate(duration = if_else(
    is.na(Reign_start) | is.na(Reign_End),
    NA_real_,                     # If end or start NA then duration NA
    Reign_End - Reign_start       # if not calculate duration
  ))

```

## Calculate the average duration of reign for all rulers
Do you remember how to calculate an average on a vector object? If not, review the last two lessons and remember that a column is basically a vector. So you need to subset your `kings` dataset to the `duration` column. If you subset it as a vector you can calculate average on it with `mean()` base-R function. If you subset it as a tibble, you can calculate average on it with `summarize()` tidyverse function. Try both ways!

* You first need to know how to select the relevant `duration` column. What are your options?
* Is your selected `duration` column a tibble or a vector? The `mean()` function can only be run on a vector. The `summarize()` function works on a tibble.
* Are you getting an error that there are characters in your column? Coerce your data to numbers with `as.numeric()`.
* Remember to handle NAs: `mean(X, na.rm=TRUE)`


```{r}
kings %>%
  summarise(mean(duration, na.rm=TRUE)) # calculates average without missing

avg_duration <- kings %>%
  summarise(avg = mean(duration, na.rm = TRUE)) %>%
  pull(avg) # now the average is saved as a value.
```

## How many and which kings enjoyed a longer-than-average duration of reign?

You have calculated the average duration above. Use it now to `filter()` the `duration` column in  `kings` dataset. Display the result and also  count the resulting rows with `count()`

```{r}
kings %>%
  filter(duration >avg_duration) %>% #filter only above average
  count() # counts how many rules are above avarage duration
```


## How many days did the three longest-ruling monarchs rule? 

* Sort kings by reign `duration` in the descending order. Select the three longest-ruling monarchs with the `slice()` function
* Use `mutate()` to create `Days` column where you calculate the total number of days they ruled
* BONUS: consider the transition year (with 366 days) in your calculation!

```{r}
# danner object med de tre monarker
top_three <- kings %>%
  arrange(desc(duration)) %>%     # arranges from high to low duration
  slice_head(n = 3)               # picks the 3 highest


# danner object med de 3 monarkers regeringstid i dage
top_three_days_simple <- top_three %>%
  mutate(Days = duration * 365L)

# Bonus version: object med de 3 monarkers regeringstid i dage + skudårsdage
# hjælpefunktion til at tælle skudår
count_leaps <- function(start_year, end_year) { # creates function
  yrs <- seq(start_year, end_year - 1)              # creates sequence counting whole years in duration
  is_leap <- (yrs %% 400 == 0) | (yrs %% 4 == 0 & yrs %% 100 != 0) # test if year is leapyear
  sum(is_leap, na.rm = TRUE) # counts leapyears 
}
# calculating for the three monarcs with leap years
top_three_days_bonus <- top_three %>%
  rowwise() %>% # does it row by row
  mutate(
    LeapDays = count_leaps(Reign_start, Reign_End), #counts leap years
    Days_adjusted = duration * 365L + LeapDays      # years in days + leap days
  ) %>%
  ungroup() # end of rowwise

select(top_three_days_bonus, Name, Days_adjusted) # shows the results
```


## Challenge: Plot the kings' duration of reign through time

What is the long-term trend in the duration of reign among Danish monarchs? How does it relate to the historical violence trends ? 

* Try to plot the duration of reign column in `ggplot` with `geom_point()` and  `geom_smooth()`
* In order to peg the duration (which is between 1-99) somewhere to the x axis with individual centuries, I recommend creating a new column `midyear` by adding to `startYear` the product of `endYear` minus the `startYear` divided by two (`startYear + (endYear-startYear)/2`).
* Now you can plot the kings dataset, plotting `midyear` along the x axis and `duration` along y axis
* BONUS: add a title, nice axis labels to the plot and make the theme B&W and font bigger to make it nice and legible!

```{r}
# creates the midyear column
kings <- kings %>%
  mutate(midyear = Reign_start + (duration / 2)      # calculates reign half way point
  )

# fjerner missing
kings_clean <- kings %>%  
  filter(!is.na(midyear), !is.na(duration)) # only keep columns where both midyear and duration is not NA

# creates plot
library(ggplot2)
ggplot(kings_clean, aes(x = midyear, y = duration)) +
  geom_point(size = 3, alpha = 0.7) + # creates points in plot
  geom_smooth(method = "loess", se = FALSE, color = "black") + # creates the line in plot
  theme_bw(base_size = 16) +   # sort/hvid og større font
  labs(
    title = "Duration of Reign Over Time",
    x = "Midyear of Reign",
    y = "Duration of Reign (years)"
  )

```

And to submit this rmarkdown, knit it into html. But first, clean up the code chunks, adjust the date, rename the author and change the `eval=FALSE` flag to `eval=TRUE` so your script actually generates an output.
Well done!


